---
title: "Quantified Self Predictions"
author: "Panos Rontogiannis"
date: "`r Sys.Date()`"
output: html_document
---

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.

## What you should submit

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).
2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details. 



```{r global_options, message=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/', warning=FALSE, message=FALSE)

library(caret)
```

## The Data

> Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.


```{r load_data, cache = TRUE}
training_orig <- read.csv('pml-training.csv.gz')
testing_final <- read.csv('pml-testing.csv.gz')
```

Initially we have `r nrow(training_orig)` observations for training and `r nrow(testing_final)` for final testing. Both datasets have `r ncol(training)` variables (columns). Let's split the training data into two more sets. Firstly the actual training dataset and secondly the testing set to be used to select the best model.

```{r split_data, cache = TRUE}
set.seed(3456)
inTrain <- createDataPartition(training_orig$classe, p=0.7, list=FALSE)
training <- training_orig[inTrain,]
testing <- training_orig[-inTrain,]
```

So now we have `r nrow(training)` observations for training our models and `r nrow(testing)` for evaluating the trained models and picking the best one.

## Exploratory Analysis

Let's explore the training and testing datasets. Let's start by looking for missing values.

```{r explor_missing_values_pt1}
complete_training <- complete.cases(training)
```

From the training data we can see that `r sum(complete_training)` out of `r nrow(training)` observations are complete (no missing values at any column). Looking more closely we can see that there are columns that mostly consist of missing values.

```{r explor_missing_values_pt2}
# Count missing values per column.
na_count <- colSums(is.na(training))
names(na_count) <- names(training)
na_count <- na_count[na_count > 0]

print(data.frame(na_count))

# Remove columns with missing values.
training_no_na <- training[, !(names(training) %in% names(na_count))]
```

After removing those `r length(na_count)` columns we are left with `r ncol(training_no_na)`.

Next let's see if there are any variables with near zero variance.

```{r explor_near_zero_var}
nzv <- nearZeroVar(training_no_na)
# Remove columns with near zero variance.
training_no_na_nzv <- training_no_na[, -nzv]
```

After removing those `r length(nzv)` columns we are left with `r ncol(training_no_na_nzv)`.

Now let's look for high correlations between the remainder varables.

```{r explor_correlations}
# Need to remove factor variables first for the cor function to work.
no_factors <- c()
for (f in colnames(training_no_na_nzv)) {
    if (!is.factor(training_no_na_nzv[, f]))
        no_factors <- c(no_factors, f)
}
cor_training <- cor(training_no_na_nzv[, no_factors])
# find highly correlated descriptors
corrDescr <- findCorrelation(cor_training, cutoff = 0.75)
# remove them from training set
training_no_na_nzv_cor <- training_no_na_nzv[, -corrDescr]
```

After removing those `r length(corrDescr)` columns we are left with `r ncol(training_no_na_nzv_cor)`.

## Training

To train the classifier we can use the following methods:

1. Regression Trees (rpart)

```{r training_rpart, cache=TRUE}
m_no_na_nzv_cor_rpart <- train(classe ~ ., data = training_no_na_nzv_cor, method='rpart')
print(m_no_na_nzv_cor_rpart)

#m_no_na_nzv_cor_rpart_pre <- train(classe ~ ., data = training_no_na_nzv_cor, method='rpart', preProcess = c('center', 'scale'))
#print(m_no_na_nzv_cor_rpart_pre)
```

2. Random Forest (rf). Though random forests are prone to overfitting, the train function will perform cross validation automatically so we don't need to worry about it. 

```{r training_rf, cache=TRUE}
m_no_na_nzv_cor_rf <- train(classe ~ ., data = training_no_na_nzv_cor, method='rf', prox=TRUE)
print(m_no_na_nzv_cor_rf)
```

3. Stochastic Gradient Boosting (gbm)

```{r training_gbm, cache=TRUE}
m_no_na_nzv_cor_gbm <- train(classe ~ ., data = training_no_na_nzv_cor, method='gbm', verbose=FALSE)
print(m_no_na_nzv_cor_gbm)
```

Now let's compare all models. Firstly we can use the resamples function:

```{r training_comparison}
resamps <- resamples(list(RPART = m_no_na_nzv_cor_rpart, RPART_PRE = m_no_na_nzv_cor_rpart_pre, RF = m_no_na_nzv_cor_rf, GBM = m_no_na_nzv_cor_gbm))
summary(resamps)
```

We can see that the random forest method is more accurate.

Finally let's use the testing dataset (taken from the original trianing set) to see how the three models perform compared to the actual classes.

```{r training_comparison_pt2}

```

## Results


## References

The data for this project come from this [source](http://groupware.les.inf.puc-rio.br/har) (see the section on the Weight Lifting Exercise Dataset).

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. [Qualitative Activity Recognition of Weight Lifting Exercises](http://groupware.les.inf.puc-rio.br/work.jsf?p1=11201). Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013. 
